Fine-scale Twitter Data Reveal Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-06-23 doc/temp-sentiment.tex

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality or no air-conditioning \cite{Samuelson2020Jun}. Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, with these heterogeneities in vulnerability being more pronounced within city regions rather than between city regions. Indeed, other studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

@@

\subsection*{Sentiment Comparison With Other Events}
We compared the expressed sentiment from exposure to temperatures to other events that have large impacts on sentiment (see Fig. \ref{fig:compare}). In particular, we examine change in sentiment from lower to higher temperatures, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F) and compare to changes in sentiment from Saturday to Monday and Hurricane Sandy the week before and after landfall. Previous work indicates sentiment expressed on Twitter are generally highest on Saturdays and lowest on Mondays, while conversely suicide rates and other metrics of mental health are lowest and highest, respectively \cite{CDC2021}. Hurricane Sandy was one of the most damaging and expensive hurricanes to make landfall in the USA in the last decade and hurricanes are associated with mental health effects including PTSD \cite{Schwartz2017Aug, Schwartz2018May}. Similar disasters have also been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

@@

% Comment on Race, why no effect on hispanics?
We found that majority Black neighborhoods are much more affected by higher temperatures than neighborhoods with a non-Black majority. This is surprising, given that Hispanics can be marginalized in both income and housing. This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than Black Americans, or may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability to heat, independently of income. This suggests that racism and patterns of housing discrimination like the legacy of redlining contribute to making Black Americans more vulnerable to heat, event at similar levels of income.

>>>>

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec} Also, while minority groups and poorer people could be expanding their livelihood activities to increase their income levels by engaging in a range of activities, exclusions based on factors like race and gender limit such opportunities (Hallegatte et al. 2015). Diverse dimensions of the factors of inequalities have implications on coping abilities and further aggravate vulnerability. Thus, particular groups and conditions, for example race/ethnicity and density (urbanization) have been identified as having differential exposure or vulnerability to extreme events (Cardona et al 2012). For instance, there is evidence for increased vulnerability to extreme events among migrant groups because of language barriers hence inability to understand extreme event-related information and also their prioritization of finding employment and housing (Thomas et al 2019). Heterogeneities in vulnerability are also more pronounced within city regions rather than between cities. In the United States, considerable regional variability is evident (Cardona et al 2012). Some studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}. So there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes. with these . Importantly, different degrees of inequality combine and interact for systems that relate directly to differential levels of social vulnerability, both in normal times and in the context of a disaster (Kuran et al 2020). This refers to the ‘concept of intersectionality’, where according to Kuran et al 2020, a factor may be a significant variable but only when allied with other variables. It primarily concerns how intersecting factors be it socioeconomic factors affect individuals who face multiple social inequities, with consequent multiple marginalisation.

@@

\subsection*{Sentiment Comparison With Other Events}
We compared the expressed sentiment from exposure to temperatures to two other events associated with impacts on sentiment and other mental health outcomes: the weekly change from Saturday to Monday, as well as the impact off a major natural disaster (see Fig. \ref{fig:compare}). For this comparison, we define heat impact as the changes in sentiment associated with temperatures increasing from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F); we define weekly changes as the difference between the mean Saturday sentiment and the mean Monday sentiment, calculated across the entire dataset; and we define the impact of Hurricane Sandy as the change in sentiment from the week before the hurricane made landfall to the week after, calculated for only impacted counties. Both of these events are associated with mental health impacts: expressed sentiment on Twitter is highest on Saturdays and lowest on Mondays, a weekly pattern that mirrors the dynamics of suicide rates \cite{CDC2021}, while Hurricane Sandy was associated with mental health effects on victims including increased anxiety, PTSD, and depression \cite{Schwartz2017Aug, Lieberman-Cribbin2017}.

@@

\subsection{Temperature Effects by Time of Day}
In addition to providing data with high spatial resolution, twitter data also comes with very high temporal resolution as each tweet is time-stamped.  Thus, we were able to examine the impact of heat on sentiment over the course of the day (See Fig. \ref{fig:ts-wbgt}).  We found that heat is associated with improved sentiment for a brief period in the late morning, and then has a consistently negative effect on sentiment during the latter half of the day, from noon until 9pm.  This effect weakens in the early evening through midnight, and then increases greatly throughout the night.  Heat has the greatest effect on expressed sentiment at 6am.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/ts_heat.png}
  \caption{Effects of rising temperatures on sentiment by hour of the day, with a 95\% confidence interval.  The value shown is the predicted change in VADER score for a 1\textdegree C WBGT increase in temperatures.}
  \label{fig:ts-wbgt}
\end{figure}

@@

Our study found disproportionate effects of climate change on identifiable and marginalized populations. We however found that majority Black neighborhoods are much more affected by higher temperatures than neighborhoods with a non-Black majority. This is surprising, given that Hispanics can be marginalized in income, language and housing. This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than Black Americans. It may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability to heat, independently of income. This suggests that racism and patterns of housing discrimination like the legacy of redlining contribute to making Black Americans more vulnerable to heat, event at similar levels of income. Thus, our study shows that, the degree of vulnerability could be related to a category of different dimensions (like combinations of race and housing) and not to a single dimension (race or income). This explains how complex vulnerability can be assessed. It further concurs with the ‘concept of intersectionality’ where the degree of vulnerability does not depend on a single dimensional attribution but results from a complex relationship between different factors such as a combination of race, housing and income (Kuran et al 2020).
